# Data import and export

To start working with data in R, you need to import it into the R environment. R supports a wide range of file formats for data import, including CSV, Excel, and databases. You can also export data from R to different file formats.

## Importing data

You can import data into R using functions like `read.csv()`, `read.table()`. Here's an example of importing a CSV file:

```{r eval=FALSE}
# Import data from a CSV file
# Daca can come from the internet or local files
data <- read.csv("https://raw.githubusercontent.com/SeviJordi/R_manual/refs/heads/main/data/iris.csv")

head(data)  # Display the first few rows of the data
```

```{r echo=FALSE}
library(knitr)
data <- read.csv("https://raw.githubusercontent.com/SeviJordi/R_manual/refs/heads/main/data/iris.csv")
kable(head(data))
```

::: callout-tip
Tabular data can be separated by commas, tabs, or other delimiters. You can specify the delimiter using the `sep` argument in the `read.csv()` function. By default, it assumes a comma as the delimiter.
:::

## Exporting data

You can export data from R to different file formats using functions like `write.csv()`, `write.table()`. Here's an example of exporting data to a CSV file:

```{r eval=FALSE}
# Export data to a CSV file
write.csv(data, "data.csv") # Save the data to a file named "data.csv"
```

By default, the `write.csv()` function includes row names in the output file. You can disable this by setting `row.names = FALSE`.

```{r eval=FALSE}
write.csv(data, "data.csv", row.names = FALSE)
```

`write.csv()` stores data in CSV format, but you can use `write.table()` to export data in other formats like tab-separated values (TSV) or plain text.

```{r eval=FALSE}
write.table(data, "data.txt", sep = "\t") # Export data as tab-separated values
```

## Excercises

::: callout-important
## Exercise

1.  Read the CSV file kleb_data.csv from the following URL: <https://raw.githubusercontent.com/SeviJordi/R_manual/refs/heads/main/data/kleb_data.csv>.

2.  Show the first rows of the data.

3.  Export the data to a tab-separated values (TSV) file named `kleb_data.tsv`.
:::

::: {.callout-note collapse="true"}
## Solution

```{r eval=FALSE}

# Exercise 1
data <- read.csv("https://raw.githubusercontent.com/SeviJordi/R_manual/refs/heads/main/data/kleb_data.csv")

# Exercise 2
head(data)

# Exercise 3
write.table(data, "kleb_data.tsv", sep = "\t")
```
:::
